Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes

### 2024-11-26 paper/appendix/bayes.tex
Updates from Overleaf

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/mh.png}
    \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2. Here, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.}
    \label{fig:mh}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts.png}
    \caption{Convergence diagnostics for a No-U-Turn Sampler (NUTS) with uniform priors on a compact set. Again, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler explores more of the posterior than the Metropolis-Hastings sampler, but fails to converge quickly.}
    \label{fig:nuts}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery_Rate.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death_Rate.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal_of_Mean_Immunity_Duration.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend_in_Force_of_Infection.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise_in_Force_of_Infection.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
    \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud. The posterior estimates from each chain are largely in agreement.}
    \label{fig:posteriors}
\end{figure}

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/mh.png}
    \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2. Here, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.}
    \label{fig:mh}
\end{figure}
\FloatBarrier
\newpage

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts.png}
    \caption{Convergence diagnostics for a No-U-Turn Sampler (NUTS) with uniform priors on a compact set. Again, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler explores more of the posterior than the Metropolis-Hastings sampler, but fails to converge quickly.}
    \label{fig:nuts}
\end{figure}
\FloatBarrier
\newpage

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery_Rate.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death_Rate.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal_of_Mean_Immunity_Duration.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend_in_Force_of_Infection.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise_in_Force_of_Infection.png}
    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
    \caption{Posterior estimates (on the logarithmic scale for everything but the trend in force of infection) from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud. The posterior estimates from each chain are largely in agreement.}
    \label{fig:posteriors}
\end{figure}

\FloatBarrier
\newpage

### 2024-11-26 paper/ms.tex
Updates from Overleaf

\arxiv{
  \usepackage{fullpage}
  %% some packages loaded by pnas-new.cls which we need for
  %% the arxiv version
  \usepackage{algorithm}
  \usepackage[noend]{algpseudocode}
  \usepackage{xcolor}
  \usepackage{tikz}
  \usepackage{enumitem}
  \bibliographystyle{ieeetr}
}{
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}
\usepackage[doublespacing]{setspace}
\usepackage[fontsize=12pt]{fontsize}}

\AtEndDocument{\refstepcounter{thm}\label{finalthm}}
\AtEndDocument{\refstepcounter{lem}\label{finallem}}
\AtEndDocument{\refstepcounter{defn}\label{finaldefn}}
\AtEndDocument{\refstepcounter{prop}\label{finalprop}}
\AtEndDocument{\refstepcounter{aspt}\label{finalaspt}}

@@

\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability.

@@

\arxiv{}{\vspace*{-3ex}}
\section{Problem Setup}

@@

\arxiv{}{\vspace*{-3ex}}
\section{Off-Parameter Particle Filters}

@@

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{Algorithm Outline}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{Discounting Weights for a Bias-Variance Tradeoff}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, naesseth18}}

@@

\begin{figure}[htbp!]
\arxiv{}{\vspace*{-2ex}}
  \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
    \arxiv{}{\vspace*{-3ex}}
    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$. \textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}. We display the MSE of score estimates for the trend in transmission at the MLE. Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates. }
    \label{fig:biasvar}
    \arxiv{}{\vspace*{-4ex}}
\end{figure}

@@

\arxiv{}{\vspace*{-3mm}}
\section{Practical Maximum-Likelihood Estimation}
\arxiv{}{\vspace*{-1mm}}
%One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.

@@

\begin{algorithm}[h!]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 and MOP-$\alpha$ cooling schedules $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run initial IF2 search under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} Set $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, and consider any $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta_m (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}

\arxiv{}{\vspace*{-1ex}}
\subsection{Optimization}

@@

\arxiv{}{\vspace*{-2ex}}
\subsection{Linear Convergence Rates}

@@

\arxiv{}{\vspace*{-3ex}}
\section{Application to a Cholera Transmission Model}
\label{sec:cholera}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{Results}

@@

\begin{figure}[htbp!]
    \centering
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/pairs.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/qq.png}
    %\arxiv{}{\vspace*{-1mm}}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. \textbf{Left:} Paired searches from the same starting point. Tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every initialization. \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. IFAD performs best on average, and approaches the MLE while no IF2 search successfully gets within 7 log-likelihood units of it. The dotted red line shows the true maximized log-likelihood. }
    \label{fig:scatter}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/2}]{imgs/095/optim.png}
    \arxiv{}{\vspace*{-1mm}}
    \caption{Optimization progress of IFAD and IF2. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. Running 60 more iterations of IF2 ultimately underperforms IFAD. We see that IFAD has better tail control and successfully reaches the MLE. 
    We use a dotted red line to display the MLE.}
    \label{fig:optim}
    \arxiv{}{\vspace*{-3ex}}
\end{figure}

\arxiv{}{\vspace*{-3ex}}
\section{Application to Bayesian Inference}
\label{sec:bayes}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}]{imgs/pmcmc/nuts_eb.png}
    \arxiv{}{\vspace*{-1mm}}
    \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains. NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

@@

\arxiv{}{\vspace*{-3ex}}
\section{Theoretical Analysis of MOP-$\alpha$}
\label{sec:thms}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{MOP-$\alpha$ Targets the Filtering Distribution}

@@

\begin{proof}
    We provide a proof sketch here, deferring the details to \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary information}. 
    When $\theta=\phi$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1$ regardless of $\alpha$, reducing to the vanilla particle filter.
    When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} \, g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} \, g^\theta_{n,k_j} \big/ g^\phi_{n,k_j})$,
    resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    Estimating the likelihood with the before-resampling estimator $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, we see that the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$. 
\end{proof}

\arxiv{}{\vspace*{-4mm}}
\subsection{MOP-$1$ Score Estimate Is Consistent}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{MOP-$\alpha$ Error, Bias and Variance}

@@

\begin{thm}[MSE and Variance of MOP-$\alpha$]
    \label{thm:mop-biasvar}
    When $\alpha\in (0,1)$ and $\theta=\phi$, define $\psi_k(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)/(1-\alpha)$. 
    There exists an $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
    \vspace*{-1ex}
    \arxiv{\begin{eqnarray}
        \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \label{eq:mop-mse}
        \\
        \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\right). \label{eq:mop-variance}
        \end{eqnarray}}{\begin{align}
        &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 \label{eq:mop-mse}\\
        &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \nonumber 
        \\
        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\hspace{-0.8ex}\right). \label{eq:mop-variance}
        \end{align}}
\end{thm}
\arxiv{}{\vspace*{-1ex}}

@@

\arxiv{}{\vspace*{-1mm}}
\subsection{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to particle approximation error, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates. 
As $\alpha$ goes to $1$, choosing $k$ appropriately, the first, third and fourth term (corresponding to the variance) increases, while the second term (corresponding to the bias) decreases. 
Likewise, the first term in the variance bound in Equation \ref{eq:mop-variance} corresponds to particle approximation error, while the second corresponds to mixing error. 
The variance increases as $\alpha$ goes to $1$.

\arxiv{}{\vspace*{-1mm}}
\subsection{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between the variances of MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$. 
The phase transition arises because even though the particle filter has forgetting properties \citep{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.

@@

\arxiv{}{\vspace*{-3ex}}
\section{Computational Efficiency}

@@

Our implementation of the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python. This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.

@@

\arxiv{}{\vspace*{-3ex}}
\section{Discussion}

@@

\arxiv{}{\vspace*{-3ex}}
\subsection{Further Work}

>>>>

\arxiv{
  \usepackage{fullpage}
  %% some packages loaded by pnas-new.cls which we need for
  %% the arxiv version
  \usepackage{algorithm}
  \usepackage[noend]{algpseudocode}
  \usepackage{xcolor}
  \usepackage{tikz}
  \usepackage{enumitem}
  \bibliographystyle{ieeetr}
}{
\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%w
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}
\usepackage[doublespacing]{setspace}
\usepackage[fontsize=12pt]{fontsize}}

\AtEndDocument{\refstepcounter{thm}\label{finalthm}}
\AtEndDocument{\refstepcounter{lem}\label{finallem}}
\AtEndDocument{\refstepcounter{defn}\label{finaldefn}}
\AtEndDocument{\refstepcounter{prop}\label{finalprop}}
\AtEndDocument{\refstepcounter{aspt}\label{finalaspt}}
\AtEndDocument{\refstepcounter{figure}\label{finalfig}}

@@

\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability. \textit{Our approach is the most significant methodological advance since \cite{ionides15}, in nearly a decade, for plug-and-play statistical inference on POMP models.}

@@

\section{Problem Setup}

@@

\section{Off-Parameter Particle Filters}

@@

\begin{figure}[htbp!]
  \centering
    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
    \arxiv{}{\vspace*{-3ex}}
    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$. \textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}. We display the MSE of score estimates for the trend in transmission at the MLE. Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates. }
    \label{fig:biasvar}
    \arxiv{}{\vspace*{-4ex}}
\end{figure}

@@

\subsection{Algorithm Outline}

@@

\subsection{Discounting Weights for a Bias-Variance Tradeoff}

@@

\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, naesseth18}}

@@

@@

\section{Practical Maximum-Likelihood Estimation}

%One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.

@@

\begin{algorithm}[htbp!]
	\caption{IFAD}
    \label{alg:ifad}
	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 and MOP-$\alpha$ cooling schedules $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
        Run initial IF2 search under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
		\textbf{While} procedure not converged: \newline
		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
		\hspace*{4mm} Set $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, and consider any $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta_m (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
		\textbf{Return} $\hat{\theta} := \theta_m.$
\end{algorithm}

\subsection{Optimization}

@@

\subsection{Linear Convergence Rates}

@@

\section{Application to a Cholera Transmission Model}
\label{sec:cholera}

@@

\subsection{Results}

@@

\begin{figure}[htbp!]
    \centering
    \vspace{1mm}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/pairs.png}
    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/qq.png}
    %\arxiv{}{\vspace*{-1mm}}
    \caption{Scatterplots depicting the performance of IFAD against that of IF2. \textbf{Left:} Paired searches from the same starting point. Tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every initialization. \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. IFAD performs best on average, and approaches the MLE while no IF2 search successfully gets within 7 log-likelihood units of it. The dotted red line shows the true maximized log-likelihood. }
    \label{fig:scatter}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \vspace{1mm}
    \includegraphics[width=\arxiv{10cm}{\textwidth/2}]{imgs/095/optim.png}
    \arxiv{}{\vspace*{-1mm}}
    \caption{Optimization progress of IFAD and IF2. The dashed orange line depicts the median warm-start given by running 40 IF2 iterations. Running 60 more iterations of IF2 ultimately underperforms IFAD. We see that IFAD has better tail control and successfully reaches the MLE. 
    We use a dotted red line to display the MLE.}
    \label{fig:optim}
    \arxiv{}{\vspace*{-3ex}}
\end{figure}

\section{Application to Bayesian Inference}
\label{sec:bayes}

@@

\begin{figure}[t!]
    \centering
    \vspace{1mm}
    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
    \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains. NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}
    \label{fig:nuts-eb}
    \arxiv{}{\vspace*{-4mm}}
\end{figure}

@@

\section{Theoretical Analysis of MOP-$\alpha$}
\label{sec:thms}

@@

\subsection{MOP-$\alpha$ Targets the Filtering Distribution}

@@

\begin{proof}
    We provide a proof sketch here, deferring the details to \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary information}. 
    When $\theta=\phi$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1$ regardless of $\alpha$, reducing to the vanilla particle filter.
    When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} \, g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} \, g^\theta_{n,k_j} \big/ g^\phi_{n,k_j})$,
    resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    Estimating the likelihood with the before-resampling estimator $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, we see that the strong consistency is a direct consequence of our result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$. 
\end{proof}

\subsection{MOP-$1$ Score Estimate Is Consistent}

@@

\subsection{MOP-$\alpha$ Error, Bias and Variance}

@@

\begin{thm}[MSE and Variance of MOP-$\alpha$]
    \label{thm:mop-biasvar}
    When $\alpha\in (0,1)$ and $\theta=\phi$, define $\psi_k(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)/(1-\alpha)$. 
    There exists an $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
    \vspace*{-1ex}
    \arxiv{\begin{eqnarray}
        \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \label{eq:mop-mse}
        \\
        \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\right). \label{eq:mop-variance}
        \end{eqnarray}}{\begin{align}
        &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 \label{eq:mop-mse}\\
        &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \nonumber 
        \\
        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\right). \label{eq:mop-variance}
        \end{align}}
\end{thm}
\arxiv{}{\vspace*{-1ex}}

@@

\subsection{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to particle approximation error, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates. 
As $\alpha$ goes to $1$, choosing $k$ appropriately, the first, third and fourth term (corresponding to the variance) increases, while the second term (corresponding to the bias) decreases. 
Likewise, the first term in the variance bound in Equation \ref{eq:mop-variance} corresponds to particle approximation error, while the second corresponds to mixing error. 
The variance increases as $\alpha$ goes to $1$.

\subsection{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between the variances of MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$. 
The phase transition arises because even though the particle filter has forgetting properties \citep{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.

@@

\section{Computational Efficiency}

@@

Our implementation of the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python. This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}. This has since spurred the development of a new open source Python library named \texttt{pypomp}\footnote{The package can be found at \href{https://pypi.org/project/pypomp/}{https://pypi.org/project/pypomp/}. The source code for this paper and original implementation can be found at \href{https://github.com/hetankevin/diffPomp}{https://github.com/hetankevin/diffPomp}.}, which incorporates the improvements in methodology and computation described within this paper.

@@

\section{Discussion}

@@

\subsection{Further Work}
